Supporting Information for: A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-08-31 Manuscript/PNAS-Adams-Submitted.Rmd
S21

where $K_1$, $K_2$, $\hat\mu_{K_1}$, $\hat\mu_{K_2}$, $\hat\sigma_{K_1}$, and $\hat\sigma_{K_2}$ are as defined above. Estimates of significance of $\hat{Z}_{12}$ may be obtained from a standard normal distribution. Typically, $\hat{Z}_{12}$ is considered a two-tailed test, however directional (one-tailed) tests may be specified should the empirical situation require it [@AdamsCollyer2016; @AdamsCollyer2019b].

>>>>

where $K_1$, $K_2$, $\hat\mu_{K_1}$, $\hat\mu_{K_2}$, $\hat\sigma_{K_1}$, and $\hat\sigma_{K_2}$ are as defined above. Estimates of significance of $\hat{Z}_{12}$ may be obtained from a standard normal distribution, or permutation. Typically, $\hat{Z}_{12}$ is considered a two-tailed test, however directional (one-tailed) tests may be specified should the empirical situation require it [@AdamsCollyer2016; @AdamsCollyer2019b]. Simulations reveal that tests based on $\hat{Z}_{12}$ have appropriate type I error rates (~0.05) and reasonable model misspecification rates (~7-12%: Supporting Information).

### 2020-08-31 Manuscript/PNAS-Adams-Submitted.tex
z12
S21

\begin{align}
  \hat{Z}_{12}=\frac{\lvert{(K_{1}-\hat\mu_{K_1})-(K_{2}-\hat\mu_{K_2})}\rvert}{\sqrt{\hat\sigma^2_{K_1}+\hat\sigma^2_{K_2}}} = \frac{\lvert Z_{K_1} - Z_{K_2}\rvert}{\sqrt{2}}
\end{align}

where \(K_1\), \(K_2\), \(\hat\mu_{K_1}\), \(\hat\mu_{K_2}\),
\(\hat\sigma_{K_1}\), and \(\hat\sigma_{K_2}\) are as defined above for
equation 2. The right side of the equation illustrates that if \(Z_K\)
has already been calculated for two sampling distributions, both have
unit variance for each of the \(Z_K\) statistics. Estimates of
significance of \(\hat{Z}_{12}\) may be obtained from a standard normal
distribution. Typically, \(\hat{Z}_{12}\) is considered a two-tailed
test, however directional (one-tailed) tests may be specified should the
empirical situation require it (36, 38).

>>>>

\begin{align}
  \hat{Z}_{12}=\frac{\lvert{(K_{1}-\hat\mu_{K_1})-(K_{2}-\hat\mu_{K_2})}\rvert}{\sqrt{\hat\sigma^2_{K_1}+\hat\sigma^2_{K_2}}} 
\end{align}

where \(K_1\), \(K_2\), \(\hat\mu_{K_1}\), \(\hat\mu_{K_2}\),
\(\hat\sigma_{K_1}\), and \(\hat\sigma_{K_2}\) are as defined above.
Estimates of significance of \(\hat{Z}_{12}\) may be obtained from a
standard normal distribution, or permutation. Typically,
\(\hat{Z}_{12}\) is considered a two-tailed test, however directional
(one-tailed) tests may be specified should the empirical situation
require it (36, 38). Simulations reveal that tests based on
\(\hat{Z}_{12}\) have appropriate type I error rates
(\textasciitilde0.05) and reasonable model misspecification rates
(\textasciitilde7-12\%: Supporting Information).

### 2020-08-31 Manuscript/SupportingInfo.Rmd
S21

>>>>

## Statistical properties of $\hat{Z}_{12}$

Here we examine the statistical properties of tests for comparing the relative strength of phylogenetic signal, using $\hat{Z}_{12}$. Simulations were performed using pure-birth trees of different sizes ($n=2^5, 2^6, \cdots, 2^{10}$), and with differing levels of phylogenetic signal ($\lambda=0.0, 0.5, \cdots, 1.0$). For each combination of $n$ and $\lambda$, 101 traits were simulated on the phylogeny under a Brownian motion model of evolution. Next, the phylogenetic signal of trait 1 was compared to the signal in the remaining 100 traits in pairwise fashion, using $\hat{Z}_{12}$. The proportion of significant results provided an estimate of the type I error or model misspecification. Specifically, type I error was evaluated when traits were simulated using no input phylogenetic signal (i.e., $\lambda = 0$). Likewise, model misspecification was evaluated for data simulated with some known, but equal, level of phylogenetic signal (i.e., $\lambda > 0$). \hfill\break

**Results:** Tests revealed that across simulation conditions, the average type I error of $\hat{Z}_{12}$ was approximately 0.05 (Fig. S21). Levels of misspecification were slightly higher, and ranged on average between 7% (when comparing traits generated on the same phylogeny: Fig. S21) to ~12% (for comparisons of traits from different trees). Both results demonstrate that comparisons of the strength of phylogenetic signal using $\hat{Z}_{12}$ do not display high levels of false positives, and are thus appropriate for comparing the degree of phylogenetic signal across traits.

```{r, echo = FALSE, out.height="50%", align= "center"}
include_graphics("fig.S21.png")  
```

\textbf{Figure S21}. Type I error rates and model misspecification for $\hat{Z}_{12}$. Type I error is found on the far-left of the plot ($\lambda = 0$).

### 2020-08-31 Manuscript/SupportingInfo.tex
S21

>>>>

\hypertarget{statistical-properties-of-hatz_12}{%
\subsection{\texorpdfstring{Statistical properties of
\(\hat{Z}_{12}\)}{Statistical properties of \textbackslash hat\{Z\}\_\{12\}}}\label{statistical-properties-of-hatz_12}}

Here we examine the statistical properties of tests for comparing the
relative strength of phylogenetic signal, using \(\hat{Z}_{12}\).
Simulations were performed using pure-birth trees of different sizes
(\(n=2^5, 2^6, \cdots, 2^{10}\)), and with differing levels of
phylogenetic signal (\(\lambda=0.0, 0.5, \cdots, 1.0\)). For each
combination of \(n\) and \(\lambda\), 101 traits were simulated on the
phylogeny under a Brownian motion model of evolution. Next, the
phylogenetic signal of trait 1 was compared to the signal in the
remaining 100 traits in pairwise fashion, using \(\hat{Z}_{12}\). The
proportion of significant results provided an estimate of the type I
error or model misspecification. Specifically, type I error was
evaluated when traits were simulated using no input phylogenetic signal
(i.e., \(\lambda = 0\)). Likewise, model misspecification was evaluated
for data simulated with some known, but equal, level of phylogenetic
signal (i.e., \(\lambda > 0\)). \hfill\break

\textbf{Results:} Tests revealed that across simulation conditions, the
average type I error of \(\hat{Z}_{12}\) was approximately 0.05 (Fig.
S21). Levels of misspecification were slightly higher, and ranged on
average between 7\% (when comparing traits generated on the same
phylogeny: Fig. S21) to \textasciitilde12\% (for comparisons of traits
from different trees). Both results demonstrate that comparisons of the
strength of phylogenetic signal using \(\hat{Z}_{12}\) do not display
high levels of false positives, and are thus appropriate for comparing
the degree of phylogenetic signal across traits.

\includegraphics[width=28.93in,height=0.5\textheight]{fig.S21}

\textbf{Figure S21}. Type I error rates and model misspecification for
\(\hat{Z}_{12}\). Type I error is found on the far-left of the plot
(\(\lambda = 0\)).
